# (APPENDIX) Appendix {-}  

# Appendix A: Restricted cubic splines {#sec:rcs-app}

A Restricted Cubic Splines (RCS) model with 3 knots $\mathbf{k} = \left(k_1, \dots, k_3\right)$ can be derived from a corresponding Cubic Splines (CS) model by forcing the curve to be linear at the extremes of the exposure distribution.

The CS model with 3 knots $\mathbf{k}$ is defined as  
\begin{equation}
\mathrm{CS}(x) = \beta_1 x + \beta_2x^2 + \beta_3x^3 + \beta_4\left(x - k_1 \right)_{+}^3 + \beta_5\left( x - k_2\right)_{+}^3 + \beta_6\left( x - k_3\right)_{+}^3
(\#eq:csa)
\end{equation}
where the `+' notation has been used ($u_+ = u$ if $u \ge 0$ and $u_+ = 0$ otherwise).

A RCS model restricts the CS function in Equation \@ref(eq:csa) to be linear before the first knot ($k_1$) and after the last knot ($k_3$). 
The first linearity constraint requires the model \@ref(eq:csa) to be linear for $x \le k_1$ 
\begin{equation*}
\mathrm{CS}(x) = \beta_1 x + \beta_2x^2 + \beta_3x^3
(\#eq:cs-k1a)
\end{equation*}

Hence, $\beta_2 = 0 \land \beta_3 = 0$.

The second linearity constraint requires the model \@ref(eq:csa) to be linear for $x \ge k_3$ 
\begin{align*}
\mathrm{CS}(x) =  & \beta_1 x + \beta_4\left( x^3 - 3x^2k_1 + 3xk_1^2- k_1^3 \right) + \beta_5\left( x^3 - 3x^2k_2 + 3xk_2^2- k_2^3 \right) + \nonumber \\
	&+ \beta_6\left( x^3 - 3x^2k_3 + 3xk_3^2- k_3^3 \right) =  \nonumber \\
	&= - \left(\beta_4k_1^3 + \beta_5k_2^3 + \beta_6k_3^3 \right) + \left( \beta_1 + 3\beta_4k_1^2 + 3\beta_5k_2^2 + 3\beta_6k_3^2\right)x +  \nonumber \\
	&- 3\left( \beta_4k_1 + \beta_5k_2 + \beta_6k_3\right)x^2 - \left( \beta_4 + \beta_5 + \beta_6\right)x^3
(\#eq:cs-k3a)
\end{align*}
 
\begin{equation*}
\begin{cases}\beta_4k_1 + \beta_5k_2 + \beta_6k_3 = 0 \\ 
\beta_4 + \beta_5 + \beta_6 = 0 \end{cases}
\begin{cases}\beta_4k_1 + \beta_5k_2 - \beta_4k_3 - \beta_5 k_3 = 0 \\ 
\beta_6 = -\beta_4 - \beta_5 \end{cases}
\begin{cases}\beta_5 = -\beta_4\frac{k_3 - k_1}{k_3 - k_2} \\ 
\beta_6 = -\beta_4 + \beta_4\frac{k_3 - k_2}{k_2 - k_1} \end{cases}
\end{equation*}

\begin{equation}
\begin{cases}\beta_5 = -\beta_4\frac{k_3 - k_1}{k_3 - k_2} \\ 
\beta_6 = -\beta_4 + \beta_4\frac{k_3 - k_1}{k_3 - k_2} \end{cases}
\begin{cases}\beta_5 = -\beta_4\frac{k_3 - k_1}{k_3 - k_2} \\ 
\beta_6 = \beta_4 \frac{k_2 - k_1}{k_3 - k_2} \end{cases}
(\#eq:systema)
\end{equation}

We can rewrite equation \@ref(eq:csa) with $\beta_2 = 0 \land \beta_3 = 0$ and equations \@ref(eq:systema)

\begin{equation}
\mathrm{RCS}(x) = \beta_1 x + \beta_4 \left[ \left( x - k_1 \right)_{+}^3 - \frac{k_3 - k_1}{k_3 - k_2} \left( x - k_2  \right)_{+}^3 + \frac{k_2 - k_1}{k_3 - k_2} \left(x - k_3 \right)_{+}^3\right]
(\#eq:rcsfina)
\end{equation}

that is a function of two variables: the quantitative exposure $x$ and a transformation of $x$.


# Appendix B: Supplementary figures

```{r p-indiv, fig.cap="Study-specific quadratic associations between coffee consumption and all-cause mortality. The relative risks are presented on a log scale using 0 cups/day as referent [@crippa2016red]."}
p_indiv
```

```{r fig-dist-vi, fig.cap="Empirical distributions for within-error terms for the study-specific linear trend in a dose--response meta-analysis between processed and red meat and bladder cancer risk [@crippa2016red]."}
dist_vi
```

```{r p-chr, fig.cap="Graphical visualization of the study-specific exposure distribution for 13 studies included in a dose--response meta-analysis between red meat consumption (g per day) and bladder cancer risk. The crosses and circles are, respectively, the referent and non-referent assigned doses of red meat consumption."}
p_chr
```

```{r p-comp-os, fig.cap="Comparison of different strategies (quadratic, spike at 0, and categorical models) in a dose--response meta-analysis of coffee consumption (cups/day) and all-cause mortality. The relative risks are presented on the log scale using 1 cup/day as referent."}
p_comp
```

```{r p-coffee-vpc, fig.cap="Variance Partition Coefficient, $\\textrm{VPC}_{ij}$, versus observed dose levels plot and LOWESS smoother for dose--response meta-analysis between coffee consumption (cusp/day) and all-cause mortality using a quadratic and meta-regression model."}
p_coffee_vpc
```


# Appendix C: Supplementary tables

```{r chr-redtab}
kable(chr_red, align = "c", booktabs = T, digits = 1, 
      col.names = c("ID", "Referent", "Min", "P25", "Median", "P75", "Max"), 
      caption = "Descriptive statistics of the assigned dose levels for 13 studies 
      included in a dose--response meta-analysis between red meat consumption (g per day) and bladder cancer risk.", linesep = "")
```


```{r AIC-pi-red}
lapply(AIC_pi_red, round, 2) %>%
  do.call("cbind", .) %>%
  `rownames<-`(apply(fpgrid, 1, function(f) paste0("(", paste(f, collapse = ", "), ")"))) %>%
  rbind(
    `Best p` = apply(fpgrid[unlist(pb_red), ], 1, function(f) paste(f, collapse = ", "))
  ) %>%
  as.data.frame() %>%
  rownames_to_column(var = "p") %>%
  kable(align = "c", booktabs = T, escape = F, longtable = T,
        caption = "AIC for the study-specific second-degree fractional polynomials with power terms specified by $p$ in a dose--response meta-analysis between red meat consumption (g per day) and bladder cancer risk. The last row reports the power term corresponding to the lowest AIC.", linesep = "") %>%
  add_header_above(c(" " = 1, "Study ID" = 13))
```


```{r blup-tab}
bi %>%
  select(id, bi.1.x:bi.2.y) %>%
  kable(align = "c", booktabs = T, escape = F, digits = 3, linesep = c(""), 
        caption = "Conditional predicted coefficients for quadratic curves in 12 studies on the association between coffee consumption (cups/day) and all-cause mortality based on a one-stage (os) and two-stage (ts) approach.", 
        col.names = c("ID", "$\\hat \\beta_{1\\textrm{os}}$", "$\\hat \\beta_{2\\textrm{os}}$",
                      "$\\hat \\beta_{1\\textrm{ts}}$", "$\\hat \\beta_{2\\textrm{ts}}$"))
```
